---
title: "| OPERATIONS MANUAL\n| \n| Strategus Data Model\n| \n| \n"
author: "Marc A. Suchard"
fontsize: 12pt
geometry: margin=1in
output:
  bookdown::html_document2:
    df_print: paged
    toc: yes
    toc_depth: 2
    toc_float: yes
    number_sections: yes
    number_tables: yes
    highlight: pygments
    css: "style.css"
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    number_sections: yes
    # citation_package: natbib
    includes:
      before_body: title.tex
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
- \newcommand{\footerDate}{`r params$date`}
- \input{header.tex}
longtable: yes
mainfont: Arial
bibliography: Protocol.bib
params:
  date: '2023-06-01'
  version: 0.0.99
subtitle: 'Version: `r params$version`'
link-citations: true
csl: bmj.csl
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(dplyr)
library(kableExtra)
```

# Access and schema

This document describes the data models for `Strategus` modules and some instructions on how to access results in these models.
We hold results on the OHDSI public database server `shinydb` that runs `postgresql`.
You can install the `postgresql` JDBC drivers; this only needs to be done once.

It is recommend to specify the environmental variable `DATABASECONNECT_JAR_FOLDER=<folder-of-your-choice>` in `.Renviron` located in the user's home directory.
In the example below, however, we set the environmental variable manually in `R` to make this document more self-contained.

```{r db, eval=FALSE}
Sys.setenv(DATABASECONNECTOR_JAR_FOLDER = ".")
DatabaseConnector::downloadJdbcDrivers(dbms = "postgresql")
```

and set-up connection details, in this case using `keyring` to hide usernames and passwords for security`

```{r connect, eval=FALSE}
# OHDSI shinydb read-only credentials
connectionDetails <- DatabaseConnector::createConnectionDetails(
  dbms = "postgresql",
  server = paste(keyring::key_get("shindyDbServer"),
                 "shinydb",
                 sep = "/"),
  user = keyring::key_get("sosFqAaUser"),
  password = keyring::key_get("sosFqAaPassword"))
```


Table \@ref(tab:schema) provides the `schema` names for SOS Challenge results.

```{r schema, echo=FALSE, warning=FALSE}
schema <- readr::read_delim(col_names = FALSE, delim = ";", trim_ws = TRUE, 
                            show_col_types = FALSE,file = "
  study1 ; ???
  study2 ; ???
  study3 ; ???
  study4 ; ???
")

tab <- kable(schema, col.names = c("SOS Challenge study", "schema"),
             linesep = "", booktabs = TRUE,
             caption = "SOS Challenge study schema")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped",
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Data models

```{r echo=FALSE}
printDataModelSpecification <- function(filename, level = "###") {

  specifications <- readr::read_csv(filename, show_col_types = FALSE) 
  
  if ("optional" %in% names(specifications)) {
    specifications <- specifications %>% filter(optional == "No")
  }
  
  tables <- split(specifications, specifications$table_name)
  
  for (table in tables) {
    header <- sprintf("%s Table %s", level, table$table_name[1])
    
    table <- table %>%
      select(Field = .data$column_name, Type = .data$data_type, Key = .data$primary_key
             # , Description = .data$description
      ) %>%
      kable(linesep = "", booktabs = TRUE, longtable = TRUE)
    
    if (knitr::is_latex_output()) {
      writeLines("")
      writeLines(header)
      
      writeLines(table %>%
                   kable_styling(latex_options = "striped", font_size = latex_table_font_size) %>%
                   column_spec(1, width = "10em") %>%
                   column_spec(2, width = "5em") %>%
                   column_spec(3, width = "3em") %>%
                   column_spec(4, width = "16em"))
    } else if (knitr::is_html_output()) {
      writeLines("")
      writeLines(header)
      
      writeLines(table %>%
                   kable_styling(bootstrap_options = "striped"))
    }
  }
}
```

```{r echo=FALSE, results="asis", warning=FALSE}
modules <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, 
                            show_col_types = FALSE,file = "
  name ; url
  Characterization ; https://raw.githubusercontent.com/OHDSI/Characterization/main/inst/settings/resultsDataModelSpecification.csv
  CohortDiagnostics ; https://raw.githubusercontent.com/OHDSI/CohortDiagnostics/main/inst/settings/resultsDataModelSpecification.csv
  CohortIncidence ; https://raw.githubusercontent.com/OHDSI/CohortIncidenceModule/master/resultsDataModelSpecification.csv
  CohortMethod ; https://raw.githubusercontent.com/OHDSI/CohortMethod/main/inst/csv/resultsDataModelSpecification.csv
  PatientLevelPrediction ; https://raw.githubusercontent.com/OHDSI/PatientLevelPrediction/main/inst/settings/resultsDataModelSpecification.csv
")

for (i in 1:nrow(modules)) {
  cat("## ", modules[i,]$name)
  printDataModelSpecification(filename = modules[i,]$url,
                              level = "###")
}
```

## Fields with minimum values

Some fields contain patient counts or fractions that are easily converted to patient counts.
To prevent identifiability, these fields are subject to a minimum value.
When the value falls below this minimum, it is replaced with the negative value of the minimum.
For example, if the minimum subject count is 5, and the actual count is 2, the value stored in the data model will be -5, which could be represented as '\<5' to the user.
Note that the value 0 is permissible, as it identifies no persons.
